
\section{Magnetism: Measuring Field from Current in a Straight Wire}

\makelabheader %(Space for student name, etc., defined in master.tex)


\bigskip
\textbf{Apparatus} 

\begin{itemize} [nosep]
\item Power supply
\item Long banana wire 
\item Stand and clamps for holding wire vertically
\item Teeny-weeny compass
\end{itemize}

\bigskip
\textbf{Activity}

Before you do anything else, verify that your teeny-weeny compass actually works.  It should point north when it's far away from any magnets or steel.  

\begin{enumerate}[labparts]

\item Make a prediction: will electric current in a wire produce a magnetic field that will deflect your compass?  If so, what direction will the magnetic field point?  (Radially inwards, towards the wire?  Away from the wire?  Circumferentially around the wire?)
\answerspace{0.5 in}

Set up your equipment to do the experiment, connecting the power supply so that positive current travels \textit{up} the vertical wire.  Crank the current up all the way so that the display reads just over 3 Amps.  

\item Hold the teeny-weeny compass very close to the vertical section of the wire, so that the outside of the compass  actually \textit{touches} the wire's rubber insulation.  Move the compass on all sides of the wire (in front, behind, etc.).  Based on your observations, describe the apparent direction of the magnetic field near the wire. \label{part_straight_wire_regular}
\answerspace{0.5 in}

\item By pointing north, your compass is pointing \textit{in the direction} of the local magnetic field.  As you look down on the vertical wire from above, does the magnetic field seem to point \textit{clockwise} or \textit{counterclockwise} around the wire? 
\answerspace{0.3 in}

\item Make a prediction: what do you think will happen to the magnetic field when you reverse the direction of the current in the vertical wire, so that current travels \textit{down} instead of \textit{up}? \label{part_straight_wire_reversed}

\bigskip
\hspace{0.5in} Prediction:

\bigskip
\hspace{0.5in} Experiment:
\bigskip

\item The direction of the magnetic field is supposed to obey the following ``right-hand rule:''
\begin{quote}\textit{If you grab a wire with your right hand, with your thumb pointing in the direction of the current, your fingers should wrap around the wire in the direction of the magnetic field.}
\end{quote}
Are the results of your experiments in parts \ref{part_straight_wire_regular} and \ref{part_straight_wire_reversed} both consistent with this right-hand rule?
\answerspace{0.3 in}

\end{enumerate}


