\appendix


\section{Treatment of Experimental Data}

\textbf{Recording Data }

When performing an experiment, record all required original observations
as soon as they are made. By \char`\"{}original observations\char`\"{}
is meant what you actually see, not quantities found by calculation.
For example, suppose you want to know the stretch of a coiled spring
as caused by an added weight. You must read a scale both before and
after the weight is added and then subtract one reading from the other
to get the desired result. The proper scientific procedure is to record
both readings as seen. Errors in calculations can be checked only
if the original readings are on record.

All data should be recorded with units. If several measurements are
made of the same physical quantity, the data should be recorded in
a table with the units reported in the column heading.

\textbf{Significant Figures} 

A laboratory worker must learn to determine how many figures in any
measurement or calculation are reliable, or \char`\"{}significant\char`\"{}
(that is, have physical meaning), and should avoid making long calculations
using figures which he/she could not possibly claim to know. \textit{All
sure figures plus one estimated figure are considered significant.}

The measured diameter of a circle, for example, might be recorded
to four significant figures, the fourth figure being in doubt, since
it is an estimated fraction of the smallest division on the measuring
apparatus. How this doubtful fourth figure affects the accuracy of
the computed area can be seen from the following example.

Assume for example that the diameter of the circle has been measured
as .526\underbar{4} cm, with the last digit being in doubt as indicated by the
line under it. When this number is squared the result will contain
eight digits, of which the last five are doubtful. Only one of the
five doubtful digits should be retained, yielding a four-digit number
as the final result.

In the sample calculation shown below, each doubtful figure has a
short line under it. Of course, each figure obtained from the use
of a doubtful figure will itself be doubtful. The result of this calculation
should be recorded as 0.2771 cm\( ^{2} \), including the doubtful
fourth figure. (The zero to the left of the decimal point is often
used to emphasize that no significant figures precede the decimal
point. This zero is not itself a significant figure.)

{\centering (.526\underbar{4} cm)\( ^{2} \) = .277\underbar{09696}
cm\( ^{2} \) = 0.277\underbar{1} cm\( ^{2} \)\par}

\textit{In multiplication and division, the rule is that a calculated
result should contain the same number of significant figures as the
least that were used in the calculation.}

\textit{In addition and subtraction, do not carry a result beyond
the first column that contains a doubtful figure.}

\textbf{Statistical Analysis} 

Any measurement is an intelligent estimation of the true value of
the quantity being measured. To arrive at a \char`\"{}best value\char`\"{}
we usually make several measurements of the same quantity and then
analyze these measurements statistically. The results of such an analysis
can be represented in several ways. Those in which we are most interested
in this course are the following:

\underbar{Mean} - The mean is the sum of a number of measurements
of a quantity divided by the number of such measurements, which is
just the arithmetic mean or the so-called average. It generally represents the best estimate of true value of the measured quantity.

\underbar{Standard Deviation} - The standard deviation (\( \sigma  \))
is a measure of the range on either side of the mean within which
approximately two-thirds of the measured values fall. For example,
if the mean is 9.75 m/s\( ^{2} \) and the standard deviation is 0.10
m/s\( ^{2} \), then approximately two-thirds of the measured values
lie within the range 9.65 m/s\( ^{2} \) to 9.85 m/s\( ^{2} \). A
customary way of expressing an experimentally determined value is:
Mean\( \pm  \)\( \sigma  \), or (9.75\( \pm  \) 0.10) m/s\( ^{2} \).
Thus, the standard deviation is an indicator of the spread in the
individual measurements, and a small \( \sigma  \) implies high precision.
Also, it means that the probability of any future measurement falling
in this range is approximately two to one. The equation for calculating
the standard deviation is\[
\sigma =\sqrt{\frac{\Sigma \left( x_{i}-\left\langle x\right\rangle \right) ^{2}}{N-1}}\]
 where \( x_{i} \) are the individual measurements, \( \left\langle x\right\rangle  \)
is the mean, and N is the total number of measurements.

\underbar{\% Difference} - Often one wishes to compare the value of
a quantity determined in the laboratory with the best known or \char`\"{}accepted
value\char`\"{} of the quantity obtained through repeated determinations
by a number of investigators. \textit{The \% difference is calculated
by subtracting the accepted value from your value, dividing by the
accepted value, and multiplying by 100.} If your value is greater
than the accepted value, the \% difference will be positive. If your
value is less than the accepted value, the \% difference will be negative.
The \% difference between two values in a case where neither is an
accepted value can be calculated by choosing one as the accepted value.
